---
title: "STAT 8700 Homework 11"
author: "Brian Detweiler"
date: "Monday, December 12th"
header-includes:
  - \usepackage{color}
  - \usepackage{xcolor}
  - \usepackage{soul}
  - \usepackage{hyperref}
  
output: pdf_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, error=FALSE}
library(rstan)
library(rjags)
source("DBDA2Eprograms/DBDA2E-utilities.R")
```

# 1. The file \texttt{planes.txt} contains 30 observations of 4 variables. You can read this into \textsf{R} using \texttt{read.table("planes.txt", header=T)}. The data is from 30 Air Force missions during the Vietnam war. The 4 variables are as follows: 
#$y$ is the the number of damaged locations of the aircraft;
#$x1$ is the type of aircraft, 0 for A4, 1, for A6; 
#$x2$ is the aircraft bomb load in tons; 
#$x3$ is the total months of aircrew experience.
#Model $y$ in JAGS using Poisson regression with a log link function. Use DIC to determine which of the three explanatory variables should be included in your model. Once you have identified the ’best’ model, use it calculate 95% prediction intervals for the amount of damage for both A4 and A6 planes, with a crew with minimal, average, and maximal experience, for a minimal, average, and maximal bombload.

```{r}
planes <- read.table('planes.txt', header=TRUE)
```


```{r}
# Log Posterior (u, v space)
#fileName <- "Assignment_10_1_a"

#modelString ="
#model {
#
  #for (j in 1:count) {
    #y[j] ~ dbin(theta[j], N[j])
    #theta[j] ~ dbeta(alpha, beta)
  #}
#
#
#
#
  #lnx <- log(alpha / beta)
  #lny <- log(alpha + beta)
#
  #alpha <- u / pow(v, 2)
  #beta <- (1 - u) / pow(v, 2)
#
  #u ~ dunif(-2, 4)
  #v ~ dunif(-5, 13)
  ##u ~ dunif(0.09, 0.22)
  ##v ~ dunif(0.08, 0.61)
#}
#"
#
#writeLines(modelString, con=fileName)
#
#basementsModel = jags.model(file=fileName, 
                             #data=list(y=basement.data$y,
                                       #N=basement.data$N, 
                                       #count=length(basement.data$N)),
                             #n.chains=4)
#
#update(basementsModel, n.iter=10000)


#basementsSamples <- coda.samples(basementsModel, n.iter=200000, variable.names=c("alpha", "beta", "theta", "y", "lnx", "lny"), thin=20)
#
#basementsSamples.M <- as.matrix(basementsSamples)

#summary(basementsSamples.M)
```

\begin{flushright}
  $\blacksquare$
\end{flushright}

\pagebreak