---
title: "STAT 8700 Homework 11"
author: "Brian Detweiler"
date: "Monday, December 12th"
header-includes:
  - \usepackage{color}
  - \usepackage{xcolor}
  - \usepackage{soul}
  - \usepackage{hyperref}
  - \usepackage[vmargin=1in,hmargin=1in]{geometry}
  - \usepackage{booktabs}
  
output: pdf_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, error=FALSE}
library(rstan)
library(rjags)
library(ggplot2)
library(reshape2)
library(dplyr)
source("DBDA2Eprograms/DBDA2E-utilities.R")
```

# 1. The file \texttt{planes.txt} contains 30 observations of 4 variables. You can read this into \textsf{R} using \texttt{read.table("planes.txt", header=T)}. The data is from 30 Air Force missions during the Vietnam war. The 4 variables are as follows: 
#$y$ is the the number of damaged locations of the aircraft;
#$x1$ is the type of aircraft, 0 for A4, 1, for A6; 
#$x2$ is the aircraft bomb load in tons; 
#$x3$ is the total months of aircrew experience.
#Model $y$ in JAGS using Poisson regression with a log link function. Use DIC to determine which of the three explanatory variables should be included in your model. Once you have identified the the best model, use it calculate 95% prediction intervals for the amount of damage for both A4 and A6 planes, with a crew with minimal, average, and maximal experience, for a minimal, average, and maximal bomb load.

```{r}
planes <- read.table('planes.txt', header=TRUE)
head(planes)
```


Let's try the zero-inflated model first:

```{r}
fileName <- "Assignment_11.1"

modelString ="
model {

  for (j in 1:count) {
    y[j] ~ dpois(theta[j] * (1 - U))
    theta[j] <- exp(beta1 * x1[j] + beta2 * x2[j] + beta3 * x3[j])
  }

  U ~ dbern(pi0)
  pi0 ~ dbeta(1, 1)

  beta1 ~ dnorm(0, 1.0E-4)
  beta2 ~ dnorm(0, 1.0E-4)
  beta3 ~ dnorm(0, 1.0E-4)

}
"

writeLines(modelString, con=fileName)

planes.model.zero.infl = jags.model(file=fileName, 
                                    data=list(y=planes$y,
                                              count=length(planes$y),
                                              x1=planes$x1,
                                              x2=planes$x2,
                                              x3=planes$x3),
                                    n.chains=4)
        
update(planes.model.zero.infl, n.iter=50000)

planes.DIC.zero.infl <- dic.samples(model = planes.model.zero.infl, n.iter = 200000, thin = 50)
planes.DIC.zero.infl

planes.samples.zero.infl <- coda.samples(model = planes.model.zero.infl, 
                                         variable.names = c("y", "beta1", "beta2", "beta3"), 
                                         n.iter = 200000, 
                                         thin = 50)

diagMCMC(planes.samples.zero.infl)

```

Now let's try a non-zero-inflated Poisson regression:

```{r}
fileName <- "Assignment_11.2"

modelString ="
model {

  for (j in 1:count) {
    y[j] ~ dpois(theta[j])
    theta[j] = exp( eta[j] )
    eta[j] = beta1 * x1[j] + beta2 * x2[j] + beta3 * x3[j]
  }

  beta1 ~ dnorm(0, 1.0E-4)
  beta2 ~ dnorm(0, 1.0E-4)
  beta3 ~ dnorm(0, 1.0E-4)

  a4.min.min ~ dpois(exp(beta2 * min(x2)  + beta3 * min(x3)))
  a4.avg.min ~ dpois(exp(beta2 * mean(x2) + beta3 * min(x3)))
  a4.max.min ~ dpois(exp(beta2 * max(x2)  + beta3 * min(x3)))
  a4.min.avg ~ dpois(exp(beta2 * min(x2)  + beta3 * mean(x3)))
  a4.avg.avg ~ dpois(exp(beta2 * mean(x2) + beta3 * mean(x3)))
  a4.max.avg ~ dpois(exp(beta2 * max(x2)  + beta3 * mean(x3)))
  a4.min.max ~ dpois(exp(beta2 * min(x2)  + beta3 * max(x3)))
  a4.avg.max ~ dpois(exp(beta2 * mean(x2) + beta3 * max(x3)))
  a4.max.max ~ dpois(exp(beta2 * max(x2)  + beta3 * max(x3)))

  a6.min.min ~ dpois(exp(beta1 + beta2 * min(x2)  + beta3 * min(x3)))
  a6.avg.min ~ dpois(exp(beta1 + beta2 * mean(x2) + beta3 * min(x3)))
  a6.max.min ~ dpois(exp(beta1 + beta2 * max(x2)  + beta3 * min(x3)))
  a6.min.avg ~ dpois(exp(beta1 + beta2 * min(x2)  + beta3 * mean(x3)))
  a6.avg.avg ~ dpois(exp(beta1 + beta2 * mean(x2) + beta3 * mean(x3)))
  a6.max.avg ~ dpois(exp(beta1 + beta2 * max(x2)  + beta3 * mean(x3)))
  a6.min.max ~ dpois(exp(beta1 + beta2 * min(x2)  + beta3 * max(x3)))
  a6.avg.max ~ dpois(exp(beta1 + beta2 * mean(x2) + beta3 * max(x3)))
  a6.max.max ~ dpois(exp(beta1 + beta2 * max(x2)  + beta3 * max(x3)))
}
"

writeLines(modelString, con=fileName)

planes.model.pois = jags.model(file=fileName, 
                               data=list(y=planes$y,
                                         count=length(planes$y),
                                         x1=planes$x1,
                                         x2=planes$x2,
                                         x3=planes$x3),
                             n.chains=4)

update(planes.model.pois, n.iter=50000)


planes.DIC.pois <- dic.samples(model = planes.model.pois, n.iter = 200000, thin = 50)

planes.DIC.pois

diffdic(planes.DIC.pois, planes.DIC.zero.infl)
```

This seems to have a slightly better fit than the zero-inflated Poisson model.

```{r}
planes.samples.pois <- coda.samples(model = planes.model.pois, 
                                    variable.names = c("beta1", 
                                                       "beta2", 
                                                       "beta3",
                                                       "a4.min.min",
                                                       "a4.avg.min",
                                                       "a4.max.min",
                                                       "a4.min.avg",
                                                       "a4.avg.avg",
                                                       "a4.max.avg",
                                                       "a4.min.max",
                                                       "a4.avg.max",
                                                       "a6.min.min",
                                                       "a6.avg.min",
                                                       "a6.max.min",
                                                       "a6.min.avg",
                                                       "a6.avg.avg",
                                                       "a6.max.avg",
                                                       "a6.min.max",
                                                       "a6.avg.max",
                                                       "a6.max.max"), 
                                    n.iter = 200000, 
                                    thin = 50)

planes.samples.pois.M <- as.matrix(planes.samples.pois[,"beta1"])
```

Our 95% prediction intervals for each situation, A4 or A6 aircraft, with minimal, average, or maximal bomb load, and minimal, average, or maximal crew experience.
```{r}
summary(planes.samples.pois)

planes.df <- as.data.frame(as.matrix(planes.samples.pois))
planes.df <- planes.df %>% select(-beta1, -beta2, -beta3)

planes.df.melt <- melt(planes.df)

ggplot(planes.df.melt, aes(x=variable, y=value)) +
  geom_boxplot() +
  coord_flip() +
  labs(x = "Aircraft.BombLoad.CrewExperience", y = "Damaged Locations")

```

\begin{flushright}
  $\blacksquare$
\end{flushright}

\pagebreak