---
title: "STAT 8700 Homework 9"
author: "Brian Detweiler"
date: "Friday, November 11th, 2016"
header-includes:
  - \usepackage{color}
  - \usepackage{xcolor}
  - \usepackage{soul}
  - \usepackage{hyperref}
  
output: pdf_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(rstan)
library(rjags)
source("DBDA2Eprograms/DBDA2E-utilities.R")
```

# 1. Repeat Assignment 8, using Stan instead of JAGS.

## 8.1. Suppose we have a population described by a Normal Distribution with 
##    known variance $\sigma^2 = 1600$ and unknown mean $\mu$. 
##    4 observations are collected from the population and the corresponding values were: 
##    940, 1040, 910, and 990.

```{r, message=FALSE}
fileName <- "Assignment_9_1_1"
data <- c(940, 1040, 910, 990)
N <- length(data)
y.bar <- mean(data)

# Write the Stan model file
modelString = 
"
data {
  int<lower=0> N; 
  real<lower=0> y[N];
}
parameters {
  real<lower=0> theta;
}
model {
	theta ~ normal(1000, 200); // Prior
  y ~ normal(theta, 40);    // Likelihood
}

"

stan.file <- paste0(fileName, ".stan")
writeLines(modelString, con=stan.file)

model.data <- list(y = data, N = N)
fit <- stan(file = stan.file, 
            data = model.data, 
            iter = 120000, 
            warmup = 2000, 
            chains = 4, 
            cores = 4, 
            open_progress = F)
```


```{r}
print(fit)

fit.mcmc <- As.mcmc.list(fit)
diagMCMC(fit.mcmc)

fit.mcmc.M <- as.matrix(fit)

hist(fit.mcmc.M[,"theta"], breaks=80, freq=F)
x <- seq(0, 1200, length = 1000)
y <- dnorm(x, mean = y.bar, sd = sqrt(1600 / N))
lines(x, y, type="l", lwd=1, col="red")

CI <- quantile(fit.mcmc.M[, "theta"], c(0.025, 0.975))
CI[[1]]
CI[[2]]

abline(v=CI[[1]], col="red")
abline(v=CI[[2]], col="red")
abline(v=y.bar, col="red")
```

\begin{flushright}
  $\blacksquare$
\end{flushright}

\pagebreak

## 9.1.2. Suppose we consider a Normal population with a variance of 16, and we collect 15 
##    observations from this population with values: 
##         26.8, 26.3, 28.3, 28.5, 26.3, 31.9, 28.5, 27.2, 20.9, 27.5, 28.0, 18.6, 22.3, 25.0, 31.5.

###     (a) If we choose a Normal(20, 25) prior, 
###         Use R to find the posterior distribution for the population mean.

```{r message=FALSE}
fileName <- "Assignment_9_1_2"

var <- 16
data <- c(26.8, 26.3, 28.3, 28.5, 26.3, 31.9, 28.5, 27.2, 20.9, 27.5, 28.0, 18.6, 22.3, 25.0, 31.5)
y.bar <- mean(data)
N <- length(data)
sigma_2 <- var

# Write the Stan model
modelString = 
"
data {
  int<lower=0> N; 
  real<lower=0> y[N];
}
parameters {
  real<lower=0, upper=100> theta;
}
model {
	theta ~ normal(20, 5);   // Prior
  y ~ normal(theta, 4);    // Likelihood
}

"


stan.file <- paste0(fileName, ".stan")
writeLines(modelString, con=stan.file)

model.data <- list(y = data, N = N)
fit <- stan(file = stan.file, 
            data = model.data, 
            iter = 120000, 
            warmup = 2000, 
            chains = 4, 
            cores = 4, 
            open_progress = F)
```

```{r}
print(fit)

fit.mcmc <- As.mcmc.list(fit)
diagMCMC(fit.mcmc)

fit.mcmc.M <- as.matrix(fit)

hist(fit.mcmc.M[,"theta"], breaks=80, freq=F)
x <- seq(20, 40, length = 1000)
y <- dnorm(x, mean = y.bar, sd = sqrt(var / N))
lines(x, y, type="l", lwd=1, col="red")

CI <- quantile(fit.mcmc.M[, "theta"], c(0.025, 0.975))
CI[[1]]
CI[[2]]

abline(v=CI[[1]], col="red")
abline(v=CI[[2]], col="red")
abline(v=y.bar, col="red")
```

\pagebreak

# 9.1.3

```{r}
fileName <- "Assignment_9_1_3"
years <- c(1976:1985)

fatal.accidents <- c(24, 25, 31, 31, 22, 21, 26, 20, 16, 22)
passenger.deaths <- c(734, 516, 754, 877, 814, 362, 764, 809, 223, 1066)
death.rate <- c(0.19, 0.12, 0.15, 0.16, 0.14, 0.06, 0.13, 0.13, 0.03, 0.15)

airline.deaths <- as.data.frame(cbind(years, fatal.accidents, passenger.deaths, death.rate))

data <- fatal.accidents
N <- length(data)

modelString = 
  "
data {
  int<lower=0> N; 
  int<lower=0> y[N];
}
parameters {
  real<lower=0> theta;
}
model {
	theta ~ gamma(1, 1);   // Prior
  y ~ poisson(theta);   // Likelihood
}
"

stan.file <- paste0(fileName, ".stan")
writeLines(modelString, con=stan.file)

model.data <- list(y = data, N = N)
fit <- stan(file = stan.file, 
            data = model.data, 
            iter = 120000, 
            warmup = 2000, 
            chains = 4, 
            cores = 4, 
            open_progress = F)

print(fit)

fit.mcmc <- As.mcmc.list(fit)
diagMCMC(fit.mcmc)

fit.mcmc.M <- as.matrix(fit)

hist(fit.mcmc.M[,"theta"], breaks=80, freq=F)
x <- seq(10, 30, length = 1000)
test <- data + 1
y <- dgamma(x, shape = sum(data) + 1, rate = N + 1)
lines(x, y, type="l", lwd=1, col="red")

CI <- quantile(fit.mcmc.M[, "theta"], c(0.025, 0.975))
CI[[1]]
CI[[2]]

abline(v=CI[[1]], col="red")
abline(v=CI[[2]], col="red")
abline(v=y.bar, col="red")
```

\begin{flushright}
  $\blacksquare$
\end{flushright}

\pagebreak

## 9.1.4

```{r}

fileName <- "Assignment_9_1_4"

years <- c(1976:1985)

fatal.accidents <- c(24, 25, 31, 31, 22, 21, 26, 20, 16, 22)
death.rate <- c(0.19, 0.12, 0.15, 0.16, 0.14, 0.06, 0.13, 0.13, 0.03, 0.15)

passenger.deaths.data <- c(734, 516, 754, 877, 814, 362, 764, 809, 223, 1066)
N <- length(passenger.deaths.data)

modelString = 
  "
data {
  int<lower=0> N;
  int<lower=0> y[N];
}
parameters {
  real<lower=0> theta;
}
model {
  theta ~ gamma(1, 1);   // Prior
  y ~ poisson(theta);   // Likelihood
}
"

stan.file <- paste0(fileName, ".stan")
writeLines(modelString, con=stan.file)

model.passenger.deaths.data <- list(y = passenger.deaths.data, N = N)

fit <- stan(file = stan.file, 
            data = model.passenger.deaths.data, 
            iter = 120000, 
            warmup = 2000, 
            chains = 4, 
            cores = 4, 
            open_progress = F)

print(fit)

fit.mcmc <- As.mcmc.list(fit)
diagMCMC(fit.mcmc)

fit.mcmc.M <- as.matrix(fit)

hist(fit.mcmc.M[,"theta"], breaks=80, freq=F)
x <- seq(200, 1200, length = 1000)

y <- dgamma(x, shape = sum(data) + 1, rate = N + 1)
lines(x, y, type="l", lwd=1, col="red")

CI <- quantile(fit.mcmc.M[, "theta"], c(0.025, 0.975))
CI[[1]]
CI[[2]]

abline(v=CI[[1]], col="red")
abline(v=CI[[2]], col="red")
abline(v=y.bar, col="red")
```

\begin{flushright}
  $\blacksquare$
\end{flushright}

\pagebreak

# 2. In Section 3.7 (starting on page 74) and in class, we discussed the analysis of a bioassay
# experiment. We simulated from the posterior distribution using a uniform prior via
# the grid method (this code is in Chapter3.R).

$$
\begin{split}
  \alpha &\sim Uniform(-5, 10)\\
  \beta &\sim Uniform(-10, 40)\\
\end{split}
$$

## Use JAGS or stan to fit this model to the data, using the following prior distribution distributions (which effectively match the one used in the grid method): Don't forget to check the model diagnostics and re-simulate if necessary. Obtain  at least 10,000 simulations per chain of $\alpha$ and $\beta$, summarize each  posterior distribution (mean, variance/sd, 95\% credible interval) and overlay the simulated values on the contour plot from Chapter3.R to verify that your simulated values seem to be from the 'correct' posterior distribution. Add a variable to your model that takes the value 1 if $\beta$ > 0 and 0 otherwise. What fraction of your simulated values of $\beta$ are greater than 0?

## (c) Add the variable LD50 to your model, where LD50 is defined on page 77. Summarize the  posterior distribution of LD50. Draw a histogram of the simulated posterior values of LD50 and compare it to the histogram in page 77 (they should be very similar).

## (d) Suppose we now wish to give 20 animals a dose level of -0.2 Simulate the posterior predictive distribution of the number deaths in the group of 20 animals. Draw a histogram of this distribution and give a 95\% credible interval.

```{r}
fileName <- "Assignment_9_2_0"

x <- c(-0.86, -0.3, -0.05, 0.73)
n <- c(5, 5, 5, 5)
y <- c(0, 1, 3, 5)
y.mod <- c(0.5, 1, 3, 4.5)

data <- cbind(y, n - y)

bivariate_normal <- function(u, v) {
  mu_U <- 0
  mu_V <- 10
  
  sigma_2_U <- 4
  sigma_2_V <- 100
  
  sigma_U <- sqrt(sigma_2_U)
  sigma_V <- sqrt(sigma_2_V)
  
  rho <- 0.5
  
  first <- 1 / (2 * pi * sigma_U * sigma_V * sqrt(1 - rho^2))
  
  second <- exp(-(1 / (2 * (1 - rho^2))) 
                * (  ((u - mu_U)^2 / sigma_2_U) 
                    + ((v - mu_V)^2 / sigma_2_V) 
                    - (2 * rho * (u - mu_U) * (v - mu_V)) / (sqrt(sigma_2_U) * sqrt(sigma_2_V))))
  rval <- first * second
  return(rval)
}

bioassay <- data.frame(cbind(x, n, y, y.mod))

inv_logit <- function(x){
  exp(x) / (1 + exp(x))
}

posterior <- function(a, b) {
  temp <- 1
	x <- bioassay$x
	y <- bioassay$y
	n <- bioassay$n
	for (i in 1: length(x)) {
	  temp <- (temp * (inv_logit(a + (b * x[i]))^y[i]) 
                  * ((1 - inv_logit(a + (b * x[i])))^(n[i] - y[i])))
  }
  
  bivariate_normal(a, b) * temp
}


N <- length(x)
C <- 10000    # this just has to be large enough to ensure all phi[i]'s < 1
              # Note that JAGS will NOT WARN you if C is too small, but will sample from a truncated distribution

ones <- rep(1, N)
model.data <- list(y=y, x=x, n=n, N=N)

modelString = 
"
model {

  for (i in 1:N) {
    line[i] <- alpha + (beta * x[i])
    p[i] <- ilogit(line[i])
    y[i] ~ dbin(p[i], n[i])
  }

  # 9.2 d.) - Could make the dosage and number of animals part of the data, but we'll keep it simple for now
  px <- alpha + (beta * -0.2)
  y_pred ~ dbin(ilogit(px), 20)

  alpha ~ dunif(-5, 10)    # Prior
  beta ~ dunif(-10, 40)    # Prior

  # Question 9.2 b.)
  # This will contain the percentage of alphas and betas > 0
  beta0 <- step(beta)
  alpha0 <- step(alpha)

  ld50 <- -alpha / beta
}
"

jags.file <- paste0(fileName, ".jags")
writeLines(modelString, con=jags.file)

# There are 3 steps to getting JAGS to generate samples from the posterior distribution. 

# Step 1 is to get all the model information into JAGS, so that it can figure out which algorithm to use to generate the MC. 

bioassayModel = jags.model(file = jags.file,
                           data = model.data,
                           n.chains = 4)

#Step 2 is to run the chains for the burn-in/warm-up period

update(bioassayModel, n.iter=2000)

#Step 3 is to runs and records the MCMC samples that we will use. 

bioassaySamples <- coda.samples(bioassayModel, n.iter=10000, variable.names=c("alpha", "beta", "beta0", "alpha0", "ld50", "y_pred"))

summary(bioassaySamples)

# Plot some diagnostics and save the plots to disk
diagMCMC(codaObject = bioassaySamples, parName=c("alpha"))
diagMCMC(codaObject = bioassaySamples, parName=c("beta"))

# To draw a histogram, we need to combine all the chains into 1 matrix
alpha.M <- as.matrix(bioassaySamples[,"alpha"])
beta.M <- as.matrix(bioassaySamples[,"beta"])
alpha0.M <- as.matrix(bioassaySamples[,"alpha0"])
beta0.M <- as.matrix(bioassaySamples[,"beta0"])
ld50.M <- as.matrix(bioassaySamples[,"ld50"])
y_pred.M <- as.matrix(bioassaySamples[,"y_pred"])

sims <- data.frame(alpha.M, beta.M)
```

```{r}
posterior_contour <- function(alpha_min, 
                              alpha_max, 
                              grid_size_alpha, 
                              beta_min, 
                              beta_max, 
                              grid_size_beta, 
                              drawlabels = TRUE) {
  
  # Generate a list of alpha values
  alpha <- seq(alpha_min, alpha_max, length = grid_size_alpha)
  
  # Generate a list of beta values
  beta <- seq(beta_min, beta_max, length = grid_size_beta)
  
  # Evaluate the posterior density and all possible combinations of alpha and beta values.  
  post.dens <- outer(alpha, beta, 'posterior')
  
  # Rescale the posterior so values are now relative to height of posterior mode. 
  scaled.dens <- post.dens / max(post.dens)
  
  # Draw contour plot. 
  contour(alpha,
          beta,
          scaled.dens, 
          levels = c(0.05, 0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85, 0.95),
          xlab="alpha",
          ylab="beta",
          drawlabels = drawlabels)
}

posterior_contour(-2, 4, 2001, -5, 25, 2001)
points(sims, xlim = c(-2, 4), ylim = c(-5, 25), pch='.', col="red")


# Here's the LD50 plot 
hist(ld50.M, breaks=100, freq=F, xlim=c(-0.6, 0.4))

# Here's the predictive results for y given 20 animals and an x_i of -0.2
hist(y_pred.M, breaks=20, freq=F)
ci <- quantile(y_pred.M, c(0.025, 0.975))
ci
abline(v=ci, col="red")

```

\begin{flushright}
  $\blacksquare$
\end{flushright}

\pagebreak

## (b) 

```{r}
fileName <- "Assignment_9_3_0"

# Read in the rat data - You may need to specify the full path to the file.
ratdata <- read.table("rats.txt", header=TRUE)

# Log Posterior (u, v space)
log.post2 <- function(u, v) {
  
  
  deaths <- ratdata$y
  rats <- ratdata$N
  alpha <- exp(u + v) / (1 + exp(u))
  beta <- exp(v) / (1 + exp(u))
  ldens <- 0
  
  # Loop over each of the 71 experiments
  for(i in 1:length(rats)) {
    
    # There is no gamma function in R - have to use Log Gamma, so the density is logged
    # This is why it's addative rather than multiplicative
    # deaths[i] is the same as y_i
    # rats[i] is the same as n_i
    ldens <- (ldens
             + (lgamma(alpha + beta) + lgamma(alpha + deaths[i]) + lgamma(beta + rats[i] - deaths[i]))
             - (lgamma(alpha) + lgamma(beta) + lgamma(alpha + beta + rats[i])))
  }

  # Return the final posterior density, which is still in logged form
  ldens - 5 / 2 * log(alpha + beta) + log(alpha) + log(beta)
}

# Just defines the size of each contour: 0.05, 0.15, 0.25, ..., 0.95
contours <- seq(0.05, 0.95, 0.1)

# Do the same steps as above, but refine the grid space
# Also, I changed the length to 200, because 2001 was slowing my computer to a crawl
u2 <- seq(-2.3, -1.3, length = 200)
v2 <- seq(1, 5, length = 200)
logdens2 <- outer(u2, v2, log.post2)
# dens2 is a 200x200 matrix of probabilities for the u2, v2 values of alpha and beta
# For instance, u2[1] = -2.3, and v2[1] = 1. dens2[1, 1] = 1.978023e-14
# This is equivalent to saying p(alpha = -2.3, beta = 1) = 1.978023e-14
dens2 <- exp(logdens2 - max(logdens2))


modelString ="
model {

  for (j in 1:count) {
    y[j] ~ dbin(theta[j], N[j])
    theta[j] ~ dbeta(alpha, beta)
  }

  y72 ~ dbin(theta72, 30)
  theta72 ~ dbeta(alpha, beta)

  lnx <- log(alpha / beta)
  lny <- log(alpha + beta)

  alpha <- u / pow(v, 2)
  beta <- (1 - u) / pow(v, 2)

  u ~ dunif(0.09, 0.22)
  v ~ dunif(0.08, 0.61)
}
"

writeLines(modelString, con="rats.jags")

#ratdata$theta.bar <- ratdata$y / ratdata$N

ratsModel = jags.model(file="rats.jags", 
                             data=list(y=ratdata$y,
                                       N=ratdata$N, 
                                       count=length(ratdata$N)),
                             n.chains=4)

update(ratsModel, n.iter=2000)


ratsSamples <- coda.samples(ratsModel, n.iter=50000, variable.names=c("alpha", "beta", "theta", "y", "lnx", "lny", "y72", "theta72"), thin=5)

ratsSamples.M <- as.matrix(ratsSamples)


diagMCMC(codaObject = ratsSamples, parName="alpha")
diagMCMC(codaObject = ratsSamples, parName="beta")
diagMCMC(codaObject = ratsSamples, parName="y72")
diagMCMC(codaObject = ratsSamples, parName="theta72")


hist(ratsSamples.M[,"lnx"], breaks=50, freq=F)
hist(ratsSamples.M[,"beta"], breaks=50, freq=F)

contour(u2, v2, dens2, levels = contours, drawlabels = FALSE, xlim=c(-2.2, -1), ylim=c(1, 5))
points(ratsSamples.M[,"lnx"], ratsSamples.M[,"lny"], col="red", pch=".")

# 9.3 b.) Simulating a 72nd experiment:
# 9.3 c.) Experiment will contain 30 rats

summary(ratsSamples.M[,"theta72"])
summary(ratsSamples.M[,"y72"])
CI <- quantile(ratsSamples.M[,"y72"], probs = c(0.025, 0.975))
CI
hist(ratsSamples.M[,"y72"], breaks=30, freq=F)
abline(v=CI, col="red")

```

\begin{flushright}
  $\blacksquare$
\end{flushright}

\pagebreak

