---
title: "STAT 8700 Final Question 3"
author: "Brian Detweiler"
date: "Thursday, December 15th"
header-includes:
  - \usepackage{color}
  - \usepackage{xcolor}
  - \usepackage{soul}
  - \usepackage{hyperref}
  
output: pdf_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
library(rjags)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, error=FALSE}
```

# 3. Consider the following data representing survival times of a random sample of 20 electrical components:

```{r}
components <- c(51, 3, 17, 13, 5, 4, 17, 1, 5, 3, 8, 22, 1, 1, 13, 8, 15, 3, 1, 13)
```

# Suppose that you cannot decide which model to use and are considering the possibility that the data could be either Gamma Distributed, or Weibull Distributed, or log-Normal Distributed.


## (a) Use DIC to decide which model is preferable.

First, we'll look at the Gamma parameterized by the mean, as shown in \cite{Gamma}. 

```{r}
fileName <- "Final_3.a"

modelString ="
model{
  for (i in 1:count) {
    y[i] ~ dgamma(sh, ra)
  }

  y.pred ~ dgamma(sh, ra)
  # parameterized by mean (m) and standard deviation (sd)
  sh <- pow(m, 2) / pow(sd, 2)
  ra <- m / pow(sd, 2)
  m ~ dunif(0,100)
  sd ~ dunif(0,100)
}
"

writeLines(modelString, con=fileName)

components.model = jags.model(file=fileName, 
                              data=list(y=components,
                                        count=length(components)),
                              n.chains=4)
        
update(components.model, n.iter=50000)

components.dic <- dic.samples(model = components.model, n.iter = 200000, thin = 50)
components.dic

components.samples <- coda.samples(model = components.model,
                                        variable.names = c("y.pred", "y", "sh", "ra", "m", "sd"), 
                                        n.iter = 200000, 
                                        thin = 50)
components.samples.M <- as.matrix(components.samples)

hist(components.samples.M[,"y.pred"], breaks=100, freq=FALSE)
```

Now, we'll look at the Gamma parameterized by the mode, as also shown in \cite{Gamma}. 

```{r}
fileName <- "Final_3.a.1"

modelString ="
model{
  for (i in 1:count) {
    y[i] ~ dgamma(sh, ra)
  }

  y.pred ~ dgamma(sh, ra)

  gmean <- sh / ra

  # parameterized by mode (m) and standard deviation (sd):
  sh <- 1 + m * ra
  ra <- ( m + sqrt( m^2 + 4 * sd^2 ) ) / ( 2 * sd^2 )


  m ~ dunif(0,100)
  sd ~ dunif(0,100)
}
"

writeLines(modelString, con=fileName)

components.model.1 = jags.model(file=fileName, 
                              data=list(y=components,
                                        count=length(components)),
                              n.chains=4)
        
update(components.model.1, n.iter=50000)

components.dic.1 <- dic.samples(model = components.model.1, n.iter = 200000, thin = 50)
components.dic.1

components.samples.1 <- coda.samples(model = components.model.1,
                                        variable.names = c("y.pred", "y", "gmean", "sh", "ra", "m", "sd"), 
                                        n.iter = 200000, 
                                        thin = 50)
components.samples.M.1 <- as.matrix(components.samples.1)
hist(components.samples.M.1[,"y.pred"], breaks=100, freq=FALSE)
```

The Gamma parameterized by the mode has a slightly better deviance score.

Let's try the Weibull.

```{r}
fileName <- "Final_3.a.2"
modelString ="
model{
  for (i in 1:count) {
    y[i] ~ dweib(a, b)
  }

  y.pred ~ dweib(a, b)

  a ~ dnorm(0, 0.0001)
  b ~ dnorm(0, 0.0001)
}
"

writeLines(modelString, con=fileName)

components.model.2 = jags.model(file=fileName, 
                              data=list(y=components,
                                        count=length(components)),
                              n.chains=4)
        
update(components.model.2, n.iter=50000)

components.dic.2 <- dic.samples(model = components.model.2, n.iter = 200000, thin = 50)
components.dic.2

components.samples.2 <- coda.samples(model = components.model.2,
                                        variable.names = c("y.pred", "y", "a", "b"), 
                                        n.iter = 200000, 
                                        thin = 50)
components.samples.M.2 <- as.matrix(components.samples.2)
hist(components.samples.M.2[,"y.pred"], breaks=100, freq=FALSE)
```

That one did just a little worse. Now we'll try the log-normal.
```{r}
fileName <- "Final_3.a.3"
modelString ="
model{
  for (i in 1:count) {
    y[i] ~ dlnorm(mu, tau)
  }

  y.pred ~ dlnorm(mu, tau)

  mu ~ dnorm(mu_0, tau_0)
  tau ~ dgamma(nu_0 / 2, nu_0 * sigma20 / 2)

  mu_0 <- 10
  tau_0 <- 25

  nu_0 <- 10
  sigma20 <- 25

  sigma2 <- 1 / tau
}
"

writeLines(modelString, con=fileName)

components.model.3 = jags.model(file=fileName, 
                              data=list(y=components,
                                        count=length(components)),
                              n.chains=4)
        
update(components.model.3, n.iter=50000)

components.dic.3 <- dic.samples(model = components.model.3, n.iter = 200000, thin = 50)
components.dic.3

components.samples.3 <- coda.samples(model = components.model.3,
                                        variable.names = c("y.pred", "y", "mu", "tau", "sigma2"), 
                                        n.iter = 200000, 
                                        thin = 50)
components.samples.M.3 <- as.matrix(components.samples.3)
hist(components.samples.M.3[,"y.pred"], breaks=100, freq=FALSE)
```

This one did way worse.

Now, we can compare all of the DIC scores and rank them.

```{r}
diffdic(components.dic, components.dic.1)
diffdic(components.dic.1, components.dic.2)
diffdic(components.dic.1, components.dic.3)
```

So it looks like the Gamma parameterized by the mode had the best fit, although Weibull did almost as well. The log-normal had the worst fit by far.

## (b) For the preferred model, calculate the 95% credible interval for the mean survival time.

```{r}
hist(components.samples.M.1[,"gmean"], breaks=100, freq=FALSE)
CI <- quantile(components.samples.M.1[,"gmean"], probs = c(0.025, 0.975))
CI
abline(v=CI, col="red")
```

The 95% credible interval for the mean survival time is (`r CI[[1]]`, `r CI[[2]]`).


## (c) For the preferred model, calculate the 95% credible interval for the survival time of the next component to be tested.

```{r}
hist(components.samples.M.1[,"y.pred"], breaks=100, freq=FALSE)
CI <- quantile(components.samples.M.1[,"y.pred"], probs = c(0.025, 0.975))
CI
abline(v=CI, col="red")
```

The 95% credible interval for the survival time of the next component to be tested is (`r CI[[1]]`, `r CI[[2]]`).

\begin{flushright}
  $\blacksquare$
\end{flushright}

\begin{thebibliography}{9}

\bibitem{Gamma}
John K. Kruschke
\textit{Gamma Likelihood Parameterized by MODE and SD}\\
Doing Bayesian Data Analysis, August 9, 2012
\href{http://doingbayesiandataanalysis.blogspot.com/2012/08/gamma-likelihood-parameterized-by-mode.html}{http://doingbayesiandataanalysis.blogspot.com/2012/08/gamma-likelihood-parameterized-by-mode.html}
\end{thebibliography}

\pagebreak